---
title: "gapmider-wrangle"
author: "Ryan"
date: "4/13/2020"
output: html_document
---

# Tidy Data Tutorial
##dplyr package

```{r}
library(tidyverse) # install tidyverse package
```

We will use the read_csv() function from the readr package (part of the tidyverse)
Note: read_csv works with local filepaths as well, you could use one from your computer.
```{r}
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')
```

Explore the dataset a bit

```{r}
head(gapminder) # show the first and last 6 rows
tail(gapminder)

head(gapminder, 12) # show the first or last x number of rows
tail(gapminder, 10)

# str() will provide a sensible description of almost anything: when in      doubt, inspect using str() on some of the recently created objects to get   some ideas about what to do next.

str(gapminder)

summary(gapminder)

library(skimr)
skim(gapminder)
```

# dplyr Basics

There are five dplyr functions that you will use to do the vast majority of data manipulations:

1. filter()
  Pick observations by their values
2. select()
  Pick variables by their names
3. mutate()
  Create new variables wtih functions of existing variables
4. summarise()
  Collapse many values down to a single summary
5. arrange()
  Reorder the rows

These can all be used in conjunction with group_by(), which changes the scope of each function from operating on the entire dataset to operating on it group-by-group.

-The first argument is a data frame.
-The subsequent arguments describe what to do with the data frame. You can refer to columns in the data frame directly without using $.
-The result is a new data frame.

# filter() 
## Subset data row-wise (by rows)

A few examples:
```{r}
filter(gapminder, lifeExp < 29)
filter(gapminder, country == "Mexico")
filter(gapminder, country %in% c("Mexico", "Peru"))
filter(gapminder, country == "Mexico", year == 2002)
newdat <- filter(gapminder, country == "Brazil", year >= 1987 & year <= 2007)
mean(newdat$lifeExp)
```

# select()
## Subset data column-wise (by column, variable)

Example:
```{r}
select(gapminder, year, country, lifeExp)
```

Or whole dataframe but remove certain rows:
```{r}
select(gapminder, -continent, -lifeExp)
```

# Use selct() and filter() together

```{r}
gap_cambodia <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)
```

However, naming them and keeping track of them gets super cumbersome, which means more time to understand what’s going on and opportunities for confusion or error.

That's where piping comes in:

# Pipe operator %>%
## The RStudio keyboard shortcut: Ctrl + Shift + M

```{r}
gapminder %>% head()
```

This pipe operator takes the thing on the left-hand-side and **pipes** it into the function call on the right-hand-side. It literally drops it in as the *first* argument

```{r}
gapminder %>% head(3)
```

**When you see %>%, think "and then"**

We can use the pipe to chain two operations together:

```{r}
gap_cambodia <- gapminder %>% #take gapminder data and then
  filter(country == "Cambodia") %>% # filter Cambodia rows and then 
  select(-continent, -lifeExp) # remove continent and lifeExp columns
```


# mutate() adds new variables

we want to know each country’s annual GDP. We can multiply pop by gdpPercap to create a new column named gdp.

```{r}
gapminder %>% 
  mutate(gdp = pop * gdpPercap)
```

Calculate the population in thousands for all Asian countries in the year 2007 and add it as a new column.


```{r}
gapminder %>% 
  filter(continent == "Asia", year == 2007) %>% 
  mutate(pop_thou = pop / 1000)
```

# groub_by() operates on groups

What if we wanted to know the total population on each continent in 2002? Answering this question requires a grouping variable.

By using group_by() we can set our grouping variable to continent and create a new column called cont_pop that will add up all country populations by their associated continents:

```{r}
gapminder %>% 
  filter(year == 2002) %>% #use gapminder data and keep only 2002 and then
  group_by(continent) %>% #group 2002 data by continent and then
  mutate(cont_pop = sum(pop)) #add each contries pop in that continent
```


What if we don’t care about the other columns and we only want each continent and their population in 2002? Here’s the next function:

# summarize() with group_by()

We want to operate on a group, but actually collapse or distill the output from that group. The summarize() function will do that for us.

```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarise(cont_pop = sum(pop)) %>% 
  ungroup()
```

So summarise() will actually only keep the columns that are grouped_by or summarized.
ungroup() removes the grouping and it’s good to get in the habit of using it after a group_by()

We can use more than one grouping variable. Let’s get total populations by continent and year.
```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarise(cont_pop = sum(pop)) %>% 
  ungroup()
```


# arrange() orders rows within columns

This is ordered alphabetically, which is cool. But let’s say we wanted to order it in ascending order for year. The dplyr function is arrange().

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarise(cont_pop = sum(pop)) %>% 
  arrange(year)
```

What is the maximum GDP per continent across all years?

My answer:
```{r}
max_gdp <- gapminder %>% 
  mutate(gdp = pop * gdpPercap) %>% 
  group_by(continent) %>% 
  summarize(max(gdp))
```
Tutorial answer:
```{r}
gapminder %>%
  mutate(gdp = pop * gdpPercap) %>%
  group_by(continent) %>%
  mutate(max_gdp = max(gdp)) %>%
  filter(gdp == max_gdp)
  
```
Both Correct

Problem ex:
1. arrange your data frame in descending order (opposite of what we’ve done). Expect that this is possible: ?arrange
2. save your data frame as a variable
3. find the maximum life expectancy for countries in Asia. What is the earliest year you encounter? The latest? Hint: you can use or base::max and dplyr::arrange()…

4. Knit your RMarkdown file, and sync it to GitHub (pull, stage, commit, push)

```{r}
yr <- gapminder %>% 
  filter(continent == "Asia") %>% #filter rows of asian countries and then
  group_by(country) %>% #group by each country and then
  filter(lifeExp == max(lifeExp)) %>% #filter each country by the max life expectency, and then
  arrange(desc(year)) #arrange rows by desceding year
```



# Joining datasets

```{r}
## read in the data. (same URL as yesterday, with co2.csv instead of gapminder.csv)
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")

## explore
co2 %>% head()
co2 %>% dim() # 12

## create new variable that is only 2007 data
gap_2007 <- gapminder %>%
  filter(year == 2007) 
gap_2007 %>% dim() # 142  

## left_join gap_2007 to co2
lj <- left_join(gap_2007, co2, by = "country")

## explore
lj %>% dim() #142
lj %>% summary() # lots of NAs in the co2_2017 columm
lj %>% View() 

## right_join gap_2007 and co2
rj <- right_join(gap_2007, co2, by = "country")

## explore
rj %>% dim() # 12
rj %>% summary()
rj %>% View() 

## full_join gap_2007 and co2
fj <- full_join(gap_2007, co2, by = "country")

## explore
fj %>% dim() # 147
fj %>% summary()
fj %>% View() 
```
