---
title: "Graphics with ggplot2"
author: "Ryan"
date: "4/10/2020"
output: html_document
---

# Learning ggplot2

Reluctantly learning ggplot2, here we gooo

Install tidyverse, which contains ggplot2

```{r}
library(tidyverse)
```

Read in data from 5 different datasets to use

```{r, echo=F}
#National Parks in California
ca <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/ca.csv") 

#Acadia National Park
acadia <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/acadia.csv")

#Southeast US National Parks
se <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/se.csv")

#2016 Visitation for all Pacific West National Parks
visit_16 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/visit_16.csv")

#All Nationally designated sites in Massachusetts
mass <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/mass.csv")
```

## Plotting

ggplot likes data in the ‘long’ format: i.e., a column for every dimension, and a row for every observation. Well structured data will save you lots of time when making figures with ggplot.

First use National Park Visitation dataset:

```{r}
head(ca)

```

use the ggplot() function and bind the plot to a specific data frame using the data argument
 ggplot(data = ca)
Then add geoms(), graphical representation of the data in the plot (points, lines, bars, etc.)

ggplot2 offers many different geoms; we will use some common ones today, including: * geom_point() for scatter plots, dot plots, etc. * geom_bar() for bar charts * geom_line() for trend lines, time-series, etc.

To add a geom to the plot use + operator. Because we have two continuous variables,
let’s use geom_point() first and then assign x and y aesthetics (aes):
