---
title: "Graphics with ggplot2"
author: "Ryan"
date: "4/10/2020"
output:
  html_document: default
  pdf_document: default
---

# Learning ggplot2

Reluctantly learning ggplot2, here we gooo

Install tidyverse, which contains ggplot2

```{r}
library(tidyverse)
```

Read in data from 5 different datasets to use

```{r}
#National Parks in California
ca <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/ca.csv") 

#Acadia National Park
acadia <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/acadia.csv")

#Southeast US National Parks
se <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/se.csv")

#2016 Visitation for all Pacific West National Parks
visit_16 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/visit_16.csv")

#All Nationally designated sites in Massachusetts
mass <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/mass.csv")
```

## Plotting

ggplot likes data in the ‘long’ format: i.e., a column for every dimension, and a row for every observation. Well structured data will save you lots of time when making figures with ggplot.

First use National Park Visitation dataset:

```{r}
head(ca)

```

use the ggplot() function and bind the plot to a specific data frame using the data argument
 ggplot(data = ca)
Then add geoms(), graphical representation of the data in the plot (points, lines, bars, etc.)

ggplot2 offers many different geoms; we will use some common ones today, including: * geom_point() for scatter plots, dot plots, etc. * geom_bar() for bar charts * geom_line() for trend lines, time-series, etc.

To add a geom to the plot use + operator. Because we have two continuous variables,
let’s use geom_point() first and then assign x and y aesthetics (aes):

```{r}
ggplot(data=ca) +
  geom_point(aes(x = year, y = visitors))
```

Notes:

1. Anything you put in the ggplot() function can be seen by any geom layers that you add (i.e., *these are universal plot settings**). This includes the x and y axis you set up in aes().
2. You can also specify aesthetics for a given geom independently of the aesthetics defined globally in the ggplot() function.
3. The + sign used to add layers must be placed at the end of each line containing a layer. If, instead, the + sign is added in the line before the other layer, ggplot2 will not add the new layer and will return an error message.

The previous figure isnt very useful because we cant tell the parks apart,  so we add color:

```{r}
ggplot(data= ca)+
  geom_point(aes(x = year, y= visitors, color= park_name))

```

# Customizing Plots

Add axis labels, title, change background with theme, remove legend title 

```{r}
ggplot(data = ca) +
    geom_point(aes(x = year, y = visitors, color = park_name)) +
    labs(x = "Year",
       y = "Visitation",
       title = "California National Park Visitation") +
    theme_classic() +
    theme(legend.title=element_blank())
```

Southeast dataset

1. Using the se dataset, make a scatterplot showing visitation to all national parks in the Southeast region with color identifying individual parks

```{r}
ggplot(data= se) +
  geom_point(aes(x = year, y = visitors, color= park_name))
```

2. Change the plot so that color indicates state.

```{r}
ggplot(data= se) +
  geom_point(aes(x = year, y = visitors, color= state))
```

3. Customize by adding your own title and theme. You can also change the text sizes and angles. Try applying a 45 degree angle to the x-axis. Use your cheatsheet!

```{r}
ggplot(data= se) +
  geom_point(aes(x = year, y = visitors, color= state)) + 
  labs(x = "Year",
       y= "Visitors",title = "Southeast parks visitation") +
  theme_classic()+
  theme(legend.title=element_blank(),
        axis.text.x = element_text(angle= 45, hjust = 1, size = 14))
  
```

# Faceting

ggplot has a special technique called _faceting_ that allows the user to split one plot into multiple plots based on data in the dataset. We will use it to make a plot of park visitation by state:

```{r}
ggplot(data=se)+
  geom_point(aes(x = year, y = visitors)) +
  facet_wrap(~ state)
```

We can now make the faceted plot by splitting further by park using park_name (within a single plot):

```{r}
ggplot(data=se) +
  geom_point(aes(x = year, y = visitors, color = park_name)) +
  facet_wrap(~ state, scales = "free")
```

# Geometric objects 
## geoms()

You can use different geoms to plot the same data. To change the geom in your plot, change the geom function that you add to ggplot(). Let’s look at a few ways of viewing the distribution of annual visitation

```{r}
ggplot(data= se) +
  geom_jitter(aes(x = year, y = visitors, color = park_name),
              width = 0.1, alpha = 0.4) +
  theme(legend.position = "none") + 
  coord_flip()
```


```{r}
ggplot(data= se, aes(x = park_name, y = visitors)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust= 1))
```

However, none of these are great for visualizing data over time. We can use geom_line() in the same way we used geom_point

```{r}
ggplot(data= se, aes(x = year, y = visitors, color = park_name)) +
  geom_line()
```

To display multiple geoms in the same plot, add multiple geom functions to ggplot():

```{r}
ggplot(data = acadia) +
  geom_point(aes(x = year, y = visitors)) +
  geom_line(aes(x = year, y = visitors)) +
  geom_smooth(aes(x = year, y = visitors)) +
  labs(title = "Acadia National Park Visitation",
       y = "Visitation",
       x = "Year") +
  theme_classic()
```

If you place mappings in a geom function, ggplot2 will treat them as local mappings for the layer. It will use these mappings to extend or overwrite the global mappings for that layer only. This makes it possible to display different aesthetics in different layers.

```{r}
ggplot(data = acadia, aes(x = year, y = visitors)) +
  geom_point() +
  geom_line() +
  geom_smooth(color = "red") +
  labs(title = "Acadia National Park Visitation",
       y = "Visitation",
       x = "Year") +
  theme_classic()
```






